\\documentclass[11pt]{article}
\\usepackage[T1]{fontenc}
\\usepackage[utf8]{inputenc}
\\usepackage{geometry}
\\usepackage{xcolor}
\\usepackage{fvextra}
\\usepackage{hyperref}
\\geometry{margin=1in}
\\DefineVerbatimEnvironment{CodeBlock}{Verbatim}{breaklines,breakanywhere,fontsize=\\small}

\begin{document}
\section*{core/cross\\_article\\_attention.py}
\textbf{Source Path}: \texttt{core/cross\\_article\\_attention.py}\par
\bigskip
\begin{CodeBlock}
"""cross_article_attention.py

Cross-article attention (frozen random observer).

This module intentionally contains ONLY the modern, import-safe primitive used by
`complete_pipeline.py`:

    CrossArticleAttention

Rationale
---------
Earlier versions of this repo contained a "MonthlyPipeline" harness here that:
- imported legacy NLI extractor APIs eagerly (fragile)
- optionally fused provenance embeddings into semantic features (methodologically risky)

That harness has been removed to avoid accidental use and import-time breakage.
If you want a monthly orchestrator, use `complete_pipeline.py`.

"""

from __future__ import annotations

from typing import Optional

import torch
import torch.nn as nn


class CrossArticleAttention(nn.Module):
    """Frozen random cross-article attention.

    This module is NOT trained. Different random seeds = different "observer"
    attention lenses over the same article-token space.

    Parameters
    ----------
    feature_dim:
        Dimensionality of article token features.
    num_heads:
        Number of attention heads.
    """

    def __init__(self, feature_dim: int, num_heads: int = 8):
        super().__init__()

        self.feature_dim = int(feature_dim)
        self.num_heads = int(num_heads)

        self.attention = nn.MultiheadAttention(
            embed_dim=self.feature_dim,
            num_heads=self.num_heads,
            batch_first=True,
        )

        # Explicitly freeze: this is an inference-only, random observer lens.
        self.eval()
        for p in self.parameters():
            p.requires_grad = False

        # Keep prints minimal but informative for long experiment logs.
        print(
            "Cross-article attention initialized (FROZEN): "
            f"feature_dim={self.feature_dim}, num_heads={self.num_heads}"
        )

    @torch.no_grad()
    def forward(self, article_tokens: torch.Tensor) -> torch.Tensor:
        """Compute attention matrix between articles.

        Parameters
        ----------
        article_tokens:
            Tensor of shape (n_articles, feature_dim).

        Returns
        -------
        Tensor of shape (n_articles, n_articles).
        Entry [i, j] = how much article i attends to article j.

        Notes
        -----
        We average attention weights across heads for visualization stability.
        """
        if article_tokens.ndim != 2:
            raise ValueError(
                f"article_tokens must be 2D (n_articles, feature_dim); got shape {tuple(article_tokens.shape)}"
            )
        if int(article_tokens.shape[1]) != self.feature_dim:
            raise ValueError(
                f"feature_dim mismatch: expected {self.feature_dim}, got {int(article_tokens.shape[1])}"
            )

        x = article_tokens.unsqueeze(0)  # (1, n_articles, feature_dim)

        _, attn_weights = self.attention(
            x,
            x,
            x,
            need_weights=True,
            average_attn_weights=True,  # average heads for visualization
        )

        return attn_weights.squeeze(0)


def set_cross_article_attention_seed(seed: Optional[int]) -> None:
    """Convenience helper for deterministic observer attention.

    If you want stable 'observer lenses' across runs, set the seed before
    constructing CrossArticleAttention.

    This helper is intentionally tiny and doesn't pull in numpy.
    """
    if seed is None:
        return
    torch.manual_seed(int(seed))


if __name__ == "__main__":
    # Minimal self-test for CrossArticleAttention only
    set_cross_article_attention_seed(0)
    X = torch.randn(10, 128)
    attn = CrossArticleAttention(feature_dim=128, num_heads=8)
    A = attn(X)
    print("Attention:", A.shape, "min/max", float(A.min()), float(A.max()))

\end{CodeBlock}
\end{document}
